Numerical Evolution from the Hamiltonian

### 2022-09-07 paper/paper.tex
Freeze before separating paper

\documentclass{iopart}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric}
\usepackage{bbm}
\expandafter\let\csname equation*\endcsname\relax
\expandafter\let\csname endequation*\endcsname\relax
\usepackage{amsmath}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\undertilde}[1]{\underset{\tilde{}}{#1}}
\newcommand{\hmatrix}{\ensuremath{%
  \begin{pmatrix}
    \mathbbm{O}&\mathbbm{I}\\-\mathbbm{I}&\mathbbm{O}
  \end{pmatrix}
}}
\newcommand{\numap}[1]{\ensuremath{%
    \left\lfloor#1\right\rceil
}}

\begin{document}
\title{Numerical Evolution from the Hamiltonian}
\author{José M L Amoreira$^1$ and Luís J M Amoreira$^2$}
\address{$^1$ Departamento de Física, Instituto Superior Técnico, Lisboa,
Portugal}
\address{$^2$ Departamento de Física, Universidade da Beira Interior, Covilhã,
Portugal}
\ead{amoreira@ubi.pt}
\begin{abstract}
  We propose a numerical method for approximate calculations of the time
  evolution of point particle systems given only the system's hamiltonian
  function and initial conditions. The method both generates and solves the
  equations of motion numerically. For demonstration purposes, a working
  im\-ple\-men\-ta\-tion (written in \texttt{python}) is described and applied
  to standard problems. The method may have some pedagogical merits but the
  numerical effort of generating the equations of motion makes it unsuitable for
  actual numerical solution of ``real'' problems with any but just a few degrees
  of freedom.
\end{abstract}
\noindent{\it Keywords\/}: Classical particle dynamics,
Hamilton's equations,
Numerical methods

\submitto{\EJP}
%\maketitle
\section{Introduction}
%-----
% 1 As equações de Hamilton de um sistema de massas pontuais formam um sistema
% de N ODEs de 1ª ordem, com a forma 
% \dot x_i= G_i(t, \tilde x)
%-----
In the hamiltonian formulation of classical mechanics~\cite{gol:1980, fw:2003},
the time evolution of physical systems is governed by Hamilton's equations
which, for point particle systems, take the form
\begin{align}\label{eq:heqs}
  \dot q_i&=\pd{H}{p_i}&
  \dot p_i&=-\pd{H}{q_i},&
  i&=1, 2, \ldots, N,
\end{align}
where $N$ is the number of degrees of freedom of the system (three times the
number of particles for three dimensional, unconstrained systems), $q_i$, $p_i$
are the canonical coordinates and conjugate momenta respectively, $H=H(t, q_i,
p_i)$ is the hamiltonian function of the system (for systems with time
independent potentials and constraints, the hamiltonian is simply their
mechanical energy function~\cite{fw:2003}) and dotted symbols denote their
total derivative with respect to time. 
Renaming the canonical coordinates and momenta as
\begin{equation}
  x_j = 
  \begin{cases}
    q_i&\quad\text{if }j\leq N\\
    p_{i-N}&\quad\text{if }j> N
  \end{cases}
  \qquad j=1, \ldots, 2N,
\end{equation}
Hamilton's equations take a more unified form
\begin{equation}\label{eq:std_ode}
  \dot x_j=G_j(t, x_1, x_2, \ldots, x_{2N}),\quad j=1, \ldots, 2N
\end{equation}
where
\begin{equation}
  G_j(t, x)=
  \begin{cases}
    \frac{\partial H}{\partial x_{j+N/2}},&\quad\text{if }j\leq N\\
    \frac{\partial H}{\partial x_{j-N/2}},&\quad\text{if }j> N.
  \end{cases}
\end{equation}
This expression can be further simplified as a matrix product:
\begin{equation}\label{eq:rhsmf}
  G_i=M_{i j}\frac{\partial H}{\partial x_j},\qquad
  \text{with }
  M=
  \begin{pmatrix}
    \mathbbm{O}&\mathbbm{I}\\
    -\mathbbm{I}&\mathbbm{O}
  \end{pmatrix},
\end{equation}
where $\mathbbm{O}$ and $\mathbbm{I}$ represent, respectively, the $N\times N$
zero and identity matrices.  Equation~\eqref{eq:std_ode} shows the Hamilton's
equations for a system of point particles with $N$ degrees of freedom as a
system of $2N$ first order ordinary differential equations (ODEs) on $2N$
variables $x_1, x_2, \ldots, x_{2N}$.

In all but a small handfull of very well known simple problems, this system of
ODEs must be solved numerically. All popular library routines for solving
ODEs\footnote{Like ODEPACK \cite{odepack} for fortran, \texttt{odeint}
  \cite{odeint} in c++ boost libraries or 
  \texttt{scipy.integrate.solve\_ivp}~\cite{scipy} for python.}
require the user to supply subprograms to compute the rhs functions $G_i$.
The ODE solver invokes these subprograms to compute estimates for the
values of the unknowns $x_i$ for arbitrary time $t$, given their values at a
given instant $t_0$.

In this work, we propose an alternative numerical method, where the user only
supplies the hamiltonian function of the system, leaving its partial derivatives
to be computed numerically as well.  This procedure is computationally more
expensive, since many evaluations of the hamiltonian function are needed in
order to estimate its partial derivatives at any moment in time, and so it
really is not suitable for the analysis of complex systems with more than just a
few coupled degreess of freedom. However, for simple systems it is very
practical and, anyway, it is an interesting concept in its own, which we haven't
seen exposed elsewhere.

%-------------------------------------------------------------------------------
\section{Numerical procedure}
Standard numerical routines for solving systems of ODEs expect the problem to be
written in the form
\begin{equation}\label{eq:numheqs}
  \frac{d\psi_i}{dt}=G_i(t, \psi_1, \psi_2, \ldots),\quad i=1, \ldots, N,
\end{equation}
where $N$ is the number of unknown functions $\psi_i(t)$ (and the number of
equations in well posed problems) and $G_i(t, \psi_1, \psi_2, \ldots)\equiv
G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
the user as callable subprograms. For instance, the python standard initial
value problem solver routine has interface
\texttt{solve\_ivp(rhs, time\_interval, psi\_0)}; here, \texttt{rhs} is the name
of a phython function to compute the values of the rhs functions $G_i$.
During its execution, the ODE solver routine repeatedly invokes these
subprograms (as many times as required by the specific integration algorithm
and/or desired accuracy) in order to compute estimates for the values the
unknown functions $\psi_i$ at given values of $t$.

For Hamilton's equations, the $\psi_i$ are the coordinates and conjugate momenta
of the system, while the $G_i$ are the partial derivatives of its hamiltonian
function (reordered and eventually symmetrized, according to
eq.~\ref{eq:rhsmf}). The usual procedure is to derive analytical expressions
for the $G_i$ as functions of $t$, the coordinates and the momenta, and then to
code those expressions as subprograms to be called by the ODE solver routine.

In the method now proposed, instead, the user only has to supply code to
compute the value of the system's hamiltonian function; its partial derivatives
(the $G_i$ in eq.~\eqref{eq:numheqs}) are estimated numerically and reordered by
another subprogram which the ODE solver calls. Note that the differentiation and
reordering of the partial derivatives can be computed by a general purpose
routine. Indeed, the gradient can even be computed also by a standard numerical
library function, if available for the chosen computing system or language.
Figure~\ref{fig:a} displays flowcharts for both methods.
\begin{figure}[htb]
  \centering
    \begin{tikzpicture}[node distance=1.0cm, font=\sffamily,scale=0.7]
      \footnotesize
      \tikzstyle{io} = [trapezium,
                        trapezium left angle=70,
                        trapezium right angle=110,
                        minimum width=1.0cm,
                        minimum height=.6cm,
                        text centered,
                        draw=black,
                        fill=blue!20]
      \tikzstyle{libcode} = [rectangle,
                             rounded corners,
                             minimum width=1.5cm,
                             text width=1.5cm,
                             minimum height=0.6cm,
                             text centered,
                             draw=black,
                             fill=orange!20]
      \tikzstyle{usercode} = [rectangle,
                             %rounded corners,
                             minimum width=1.5cm,
                             minimum height=0.6cm,
                             text centered,
                             draw=black,
                             text width=1.5cm,
                             fill=green!20]
      \tikzstyle{arrow} = [thick,->,>=stealth]

      \begin{scope}[xshift=-3.5cm]
        \node at (0,1) {Standard method};
        \node (input) [io] {Initial state};
        \node (odesolver) [libcode, below of=input] {ODE Solver};
        \node (output) [io, below of=odesolver] {Final state};
        \node (rhs) [usercode, right of=odesolver, node distance=2.3cm]
          {$M\,\text{grad}\,H$};
        \draw [arrow] (input) -- (odesolver);
        \draw [arrow] (odesolver) -- (output);
        \draw [arrow] (odesolver) -- (rhs);
        \draw [arrow]  (rhs) -- (odesolver);
      \end{scope}

      \begin{scope}[xshift=3.5cm]
        \node at (0,1) {Proposed method};
        \node (input) [io] {Initial state};
        \node (odesolver) [libcode, below of=input] {ODE Solver};
        \node (output) [io, below of=odesolver] {Final state};
        \draw [arrow] (input) -- (odesolver);
        \draw [arrow] (odesolver) -- (output);
        \node (rhs) [libcode, right of=odesolver, node distance=2.5cm]
          {General purpose\\ $M\,\text{grad}\,f$ routine};
        \draw [arrow] (odesolver) -- (rhs);
        \draw [arrow]  (rhs) -- (odesolver);
        \node (h) [usercode, right of=rhs, node distance=2.5cm] {$H$ function};
        \draw [arrow] (rhs) -- (h);
        \draw [arrow] (h) -- (rhs);
      \end{scope}
    \end{tikzpicture}
  \caption{Flowcharts for the standard and the proposed methods. Round cornered
    boxes represent library or general purpose code; right angled boxes contain
    code for the specific calculation at hand, which the user must
  supply.\label{fig:a}}
\end{figure}

%To make this discussion clearer, let us consider a specific problem, the
%classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
%the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
%a classical oscilator with mass $m$ and restoring force constant $k$ is 
%\begin{equation}
%  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
%\end{equation}
%Hamilton's equations~\eqref{eq:heqs} then read
%\begin{align}
%  \frac{dq}{dt}&=\frac{p}{m}&
%  \frac{dp}{dt}&=-kq,
%\end{align}
%wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
%$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
%by iterating the formulas
%\begin{align}
%  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
%  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
%\end{align}
%where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
%implements this method, and this the usual procedure. With this paper, we
%consider a method where explicit expressions for the hamiltonian's partial
%derivatives in the equations of motion are replaced by numerical estimates. In
%our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
%and~\eqref{eq:efe2} by
%\begin{align*}
%  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
%  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
%\end{align*}
%where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
%In order to make numerical aproximations to the partial derivatives of a
%function, one must evaluate it several times; at the very least, two evaluations
%are needed. For example, using the second order central difference method, the
%derivative is estimated as
%\begin{equation}\label{eq:cdiff}
%  \frac{\partial\psi}{\partial x} = 
%    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
%  o(\delta x)^2,
%\end{equation}
%leading to approximation formulas that, for our example problem, read 
%\begin{align}
%  \numap{\frac{\partial H}{\partial p}}_i &=
%  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
%  \numap{\frac{\partial H}{\partial q}}_i &=
%  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
%\end{align}
%where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
%Other approximations have higher accuracy, at the cost of extra evaluations of
%the function to be derived.

%This need for repeated evaluation of the system's hamiltonian reduces the
%efficiency of this method. However, programming a procedure for the hamiltonian
%function is usually much simpler then doing it for the set of its partial
%derivatives.  Also, adjusting the syntax and semantics of the ODE solver and the
%hamiltonian procedure is likewise much simpler than when the partial derivatives
%are coded.  For small systems, using the hamiltonian is much simpler and faster.
%
%
%
\section{Simple implementation in Python and examples}
We made a simple implementation of this method in Python, using the standard
numerical extensions library NumPy~\cite{numpy:2011,numpy:2020}.  It consists of
a simple function that calls the hamiltonian function supplied by the user to
evaluate its partial derivatives, using the central difference formulas of
eqs.~\eqref{eq:cdiff}. This function is then supplied to NumPy's ODE solver
\texttt{solve\_ivp}.  For greater efficiency, it is wrapped in a class, whose
objects store initialization details on creation, like the hamiltonian's
parameters, the number of degrees of freedom, the values of the discretization
steps, etc. These objects are callable with signature adjusted to what the ODE
solver expects.  The whole class definition fits in a few dozen lines of code
and is very easy to use. It is available for download as free software at
github~\cite{nevh:2020}. The repository also includes a \texttt{jupyter
notebook} showcasing several example applications.

\section{Conclusion}

\section*{References}
\begin{thebibliography}{99}
  \bibitem{gol:1980} H. Goldstein, \textsl{Classical Mechanics.} Addison-Wesley
  (1980)
%
  \bibitem{fw:2003} 
    A.L.~Fetter, J.D.~Walecka,
    \textsl{Theoretical Mechanics of Particles and Continua.}
    Dover (2003)
%
  \bibitem{odepack}
  A.C. Hindmarsh, \textsl{ODEPACK, A Systematized Collection of ODE Solvers,} in
    Scientific Com\-puting, R. S. Stepleman et al. (eds.), North-Holland,
    Amsterdam, 1983 (vol. 1 of IMACS Transactions on Scientific Computation),
    pp. 55-64. 
%
  \bibitem{odeint}
  K. Ahnert and M. Mulansky, \textsl{Odeint - Solving Ordinary Differential
  Equations in C++,} AIP Conf. Proc. 1389, pp. 1586-1589 (2011);
doi:http://dx.doi.org/10.1063/1.3637934
%
  \bibitem{scipy}
  P. Virtanen et al, \textsl{SciPy 1.0: Fundamental Algorithms for Scientific
  Computing in Python}, Nature Methods 17, pp. 261--272 (2020)
%
  \bibitem{french:71} A.P.~French, \textsl{Vibration and Waves.} W.W. Norton \&
  Company (1971)
%
  \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in
  Physics.} Springer (2012)
%
  \bibitem{nr:2007} W.H.~Press, S.A.~Teukolsky, W.T.~Vetterling, B.P.~Flannery,
  \textsl{Numerical Recipes -- The Art of Scientific Computing,} 3rd Ed.
  Cambridge University Press (2007)
%
  \bibitem{numpy:2011} S.J.~van der Walt, S.C.~Colbert, G.~Varoquaux,
  \textsl{The NumPy array: a structure for efficient numerical computation,}
  Computing in Science \& Engineering, \textbf{13} no. 2, 22-30 (2011).
  doi: 10.1109/MCSE.2011.37
%
  \bibitem{numpy:2020} C.R.~Harris, K.J.~Millman, S.J.~van der Walt et. al.,
  \textsl{Array programming with NumPy,} Nature \textbf{585}, 357-362 (2020).
  https://doi.org/10.1038/s41586-020-2649-2
%
  \bibitem{nevh:2020} J.M.L.~Amoreira, L.J.M.~Amoreira,
  \texttt{https://github.com/ljmamoreira/nevh} (2020)

>>>>

\documentclass{iopart}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric}
\usepackage{bbm}
\usepackage{listings, algpseudocode, amssymb}
%\usepackage{xcolor}
\lstset{%
  language=python,
  %backgroundcolor=\color{black!5},
  basicstyle=\sffamily\footnotesize,
  keepspaces=false,
  showspaces=false,
  showstringspaces=false
}
\expandafter\let\csname equation*\endcsname\relax
\expandafter\let\csname endequation*\endcsname\relax
\usepackage{amsmath}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\undertilde}[1]{\underset{\tilde{}}{#1}}
\newcommand{\hmatrix}{\ensuremath{%
  \begin{pmatrix}
    \mathbbm{O}&\mathbbm{I}\\-\mathbbm{I}&\mathbbm{O}
  \end{pmatrix}
}}
\newcommand{\numap}[1]{\ensuremath{%
    \left\lfloor#1\right\rceil
}}

\begin{document}
\title{Numerical Evolution from the Hamiltonian}
\author{José M L Amoreira$^1$ and Luís J M Amoreira$^2$}
\address{$^1$ Departamento de Física, Instituto Superior Técnico, Lisboa,
Portugal}
\address{$^2$ Departamento de Física, Universidade da Beira Interior, Covilhã,
Portugal}
\ead{amoreira@ubi.pt}
\begin{abstract}
  We propose a numerical method for approximate calculations of the time
  evolution of point particle systems given only the system's Hamiltonian
  function and initial conditions. The method both generates and solves the
  equations of motion numerically. For demonstration purposes, a working
  im\-ple\-men\-ta\-tion written in Python is described and applied
  to standard problems. The method may have some pedagogical merits but the
  numerical effort of generating the equations of motion makes it unsuitable for
  actual numerical solution of ``real'' problems with any but just a few degrees
  of freedom.
\end{abstract}
\noindent{\it Keywords\/}: Classical particle dynamics,
Hamilton's equations,
Numerical methods

\submitto{\EJP}
%\maketitle
\section{Introduction}
%-----
% 1 As equações de Hamilton de um sistema de massas pontuais formam um sistema
% de N ODEs de 1ª ordem, com a forma 
% \dot x_i= G_i(t, \tilde x)
%-----
In the Hamiltonian formulation of classical mechanics~\cite{gol:1980, fw:2003},
the time evolution of physical systems is governed by Hamilton's equations
which, for point particle systems, take the form
\begin{align}\label{eq:heqs}
  \dot q_i&=\pd{H}{p_i}&
  \dot p_i&=-\pd{H}{q_i},&
  i&=1, 2, \ldots, N,
\end{align}
where $N$ is the number of degrees of freedom of the system (three times the
number of particles for three dimensional, unconstrained systems), $q_i$, $p_i$
are the canonical coordinates and conjugate momenta respectively, $H=H(t, q_i,
p_i)$ is the Hamiltonian function of the system (for systems with time
independent potentials and constraints, the Hamiltonian is simply the
mechanical energy function~\cite{fw:2003}) and dotted symbols denote their
total derivatives with respect to time. 
Renaming the canonical coordinates $q_i$ and momenta $p_i$ as
\begin{equation}\label{eq:xinot}
  \nu_i = 
  \begin{cases}
    q_i&\quad\text{if }i\leq N\\
    p_{i-N}&\quad\text{if }i> N
  \end{cases}
  \qquad i=1, \ldots, 2N,
\end{equation}
Hamilton's equations take a more unified form
\begin{equation}\label{eq:std_ode}
  \dot \nu_i=G_i(t, \nu_1, \nu_2, \ldots, \nu_{2N}), \quad i=1, \ldots, 2N
\end{equation}
where
\begin{equation}
  G_i(t, \nu)=
  \begin{cases}\displaystyle
    \pd{H}{\nu_{i+N}} ,&\quad\text{if }i\leq N\\[1em]
    \displaystyle
    -\pd{H}{\nu_{i-N}}, &\quad\text{if }i> N.
  \end{cases}
\end{equation}
This expression can be further shortened as a matrix product:
\begin{equation}\label{eq:rhsmf}
  G_i= \sum_j M_{i j}\pd{H}{\nu_j},\qquad
  \text{with }
  M=
  \begin{pmatrix}
    \mathbbm{O}&\mathbbm{I}\\
    -\mathbbm{I}&\mathbbm{O}
  \end{pmatrix},
\end{equation}
where $\mathbbm{O}$ and $\mathbbm{I}$ represent, respectively, the $N\times N$
zero and identity matrices.  Equation~\eqref{eq:std_ode} shows the Hamilton's
equations for a system of point particles with $N$ degrees of freedom as a
system of $2N$ first order ordinary differential equations (ODEs) on $2N$
variables $\nu_1, \nu_2, \ldots, \nu_{2N}$.

In all but a small handfull of very well known simple problems, this system of
ODEs has no analytical solutions and must be solved numerically.  Popular
computer library routines for solving ODEs\footnote{Like ODEPACK~\cite{odepack}
  for fortran, \texttt{odeint} \cite{odeint} in c++ boost libraries or
\texttt{solve\_ivp} in SciPy~\cite{scipy} for Python.} require the user to
supply subprograms to compute the right hand side functions $G_i$.  The ODE
solver invokes these subprograms to compute estimates for the values of the
unknowns $\nu_i$ for arbitrary time $t$, given their values at a particular
instant $t_0$.

In this work, we propose an alternative numerical method (which we will refer to
as \emph{Numerical Evolution from the Hamiltonian}, or NEVH) where the user only
needs to write code for the Hamiltonian function of the system, leaving its
partial derivatives to be computed numerically by a general purpose subprogram.
Figure~\ref{fig:a} displays flowcharts for both methods.
\begin{figure}[htb]
  \centering
    \begin{tikzpicture}[node distance=1.0cm, font=\sffamily,scale=0.7]
      \footnotesize
      \tikzstyle{libcode} = [rectangle,
                             rounded corners,
                             minimum width=1.5cm,
                             text width=1.5cm,
                             minimum height=0.6cm,
                             text centered,
                             draw=black,
                             %fill=orange!20
                             ]
      \tikzstyle{usercode} = [rectangle,
                             %rounded corners,
                             minimum width=1.5cm,
                             minimum height=0.6cm,
                             text centered,
                             draw=black,
                             text width=1.5cm,
                             %fill=green!20
                             ]
      \tikzstyle{arrow} = [thick,->,>=stealth]

      \begin{scope}[xshift=-3.5cm]
        \node at (0,1) {Standard method};
        \node (input) [usercode] {Initial state};
        \node (odesolver) [libcode, below of=input] {ODE Solver};
        \node (output) [usercode, below of=odesolver] {Final state};
        \node (rhs) [usercode, right of=odesolver, node distance=2.3cm]
          {Subprogram for $G_i$};
        \draw [arrow] (input) -- (odesolver);
        \draw [arrow] (odesolver) -- (output);
        \draw [arrow] (odesolver) -- (rhs);
        \draw [arrow]  (rhs) -- (odesolver);
      \end{scope}

      \begin{scope}[xshift=3.5cm]
        \node at (0,1) {NEVH};
        \node (input) [usercode] {Initial state};
        \node (odesolver) [libcode, below of=input] {ODE Solver};
        \node (output) [usercode, below of=odesolver] {Final state};
        \draw [arrow] (input) -- (odesolver);
        \draw [arrow] (odesolver) -- (output);
        \node (rhs) [libcode,  right of=odesolver,
                     node distance=2.5cm]
                     {General purpose\\ $M_{ij}%\,
                     \partial/\partial x_j$ subprogram};
        \draw [arrow] (odesolver) -- (rhs);
        \draw [arrow]  (rhs) -- (odesolver);
        \node (h) [usercode, right of=rhs, node distance=2.5cm] {$H$ function};
        \draw [arrow] (rhs) -- (h);
        \draw [arrow] (h) -- (rhs);
      \end{scope}
    \end{tikzpicture}
  \caption{Flowcharts for the standard and the proposed methods. Round cornered
    boxes represent library or general purpose code; right angled boxes contain
    code for the specific calculation at hand, which the user must
  supply.\label{fig:a}}
\end{figure}

This method is computationally more expensive the the standard approach, since
at least two evaluations of the Hamiltonian function are needed to evaluate each
of its partial derivatives at any moment in time, and so it really is not
suitable for the analysis of complex systems with more than just a few coupled
degrees of freedom. However, for simple systems it is very practical
and, anyway, it is an interesting approach in its own, which we haven't seen
exposed elsewhere.

%-------------------------------------------------------------------------------
\section{Numerical procedure and a simple implementation in Python}
The ``Generic purpose $M_{ij}\partial/\partial x_j$ subprogram'' in the
flowchart in Figure~\ref{fig:a} can be easily implemented in any modern computer
language. Using a central difference formula for the partial derivatives, a
pseudocode version can be sketched as
\begin{algorithmic}[H]
  \Function{G}{$H, t, \nu$}
  \For{$1\leq i \leq 2N$}
  \State
  $\partial_iH\gets \frac{\displaystyle%
  H(t, \tilde\nu_i,\nu_i+\delta\nu_i)-
H(t, \tilde\nu_i,\nu_i-\delta\nu_i)}{\displaystyle2\delta\nu_i}
  $
  \EndFor\State
  \Return $\{\partial_{N+1}H, \partial_{N+2}H, \ldots, \partial_{2N}H,
  -\partial_1H, \partial_2H, \ldots, \partial_NH\}$
  \EndFunction
\end{algorithmic}
(In this pseudocode definition, the notation $\tilde\nu_i$ stands for all the
variables $\nu_1, \ldots, \nu_{2N}$, \emph{except} the $i$-th, $\nu_i$.)

%Standard numerical routines for solving systems of ODEs expect the problem to be
%written in the form
%\begin{equation}\label{eq:numheqs}
%  \frac{d\nu_i}{dt}=G_i(t, \nu_1, \nu_2, \ldots, \nu_{2N}),\quad i=1, \ldots, N,
%\end{equation}
%where $N$ is the number of unknown functions $\nu_i(t)$ (and the number of
%equations in well posed problems) and $G_i(t, \nu_1, \nu_2, \ldots)\equiv
%G_i(t,\nu_j)$ are the rhs of the differential equations, which must be coded by
%the user as callable subprograms. For instance, the Python standard initial
%value problem solver routine has interface
%\texttt{solve\_ivp(rhs, time\_interval, psi\_0)}; here, \texttt{rhs} is the name
%of a phython function to compute the values of the rhs functions $G_i$.
%During its execution, the ODE solver routine repeatedly invokes these
%subprograms (as many times as required by the specific integration algorithm
%and/or desired accuracy) in order to compute estimates for the values the
%unknown functions $\nu_i$ at given values of $t$.
%
%For Hamilton's equations, the $\nu_i$ are the coordinates and conjugate momenta
%of the system, while the $G_i$ are the partial derivatives of its Hamiltonian
%function (reordered and eventually multiplied by -1, according
%to~(\ref{eq:rhsmf})). The usual procedure is to derive analytical expressions
%for the $G_i$ as functions of $t$, the coordinates and the momenta, and then to
%code those expressions as subprograms to be called by the ODE solver routine.
%Quite often, this derivation is cumbersome, time consuming and error prone.
%However, for simple systems at least, numerical calculation of the Hamiltonian
%function may be a practical alternative.
%
%In the method now proposed, instead, the user only has to supply code to
%compute the system's Hamiltonian; the rhs functions $G_i$ in~\eqref{eq:std_ode},
%needed by the ODE solver, are computed numerically by a generic subroutine
%which takes the Hamiltonian function as an argument and returns the values of
%its partial derivatives, properly reordered and multiplied by $\pm1$ as
%prescribed in~\eqref{eq:rhsmf}. This function is generic in the sense that it
%can be written independently of the particular system under analysis or the
%specific form of its Hamiltonian. For example, using central difference formulas
%and keeping the notation of~\eqref{eq:xinot}, it can be defined as 
%\begin{algorithmic}[H]
%  \Function{G}{$H, t, \nu$}
%  \For{$1\leq i \leq 2N$}
%  \State
%  $\partial_iH\gets \frac{\displaystyle%
%  H(t, \tilde\nu,\nu_i+\delta\nu_i)-
%H(t, \tilde\nu,\nu_i-\delta\nu_i)}{\displaystyle2\delta\nu_i}
%  $
%  \EndFor\State
%  \Return $\{\partial_{N+1}H, \partial_{N+2}H, \ldots, \partial_{2N}H,
%  -\partial_1H, \partial_2H, \ldots, \partial_NH\}$
%  \EndFunction
%\end{algorithmic}
%(In this pseudocode definition, the notation $\tilde\nu$ stands for all the
%variables $\nu_1, \ldots, \nu_{2N}$, except the $i$-th, $\nu_i$.)
%In the method now proposed, instead, the user only has to supply code to
%compute the value of the system's Hamiltonian function; its partial derivatives
%(the $G_i$ in~\eqref{eq:numheqs}) are estimated numerically and reordered by
%another subprogram which the ODE solver calls. Note that the differentiation and
%reordering of the partial derivatives can be computed by a general purpose
%routine. Indeed, the gradient can even be computed also by a standard numerical
%library function, if available for the chosen computing system or language.

We made a simple implementation of this method in Python, using the standard
packages NumPy~\cite{numpy:2011,numpy:2020} and SciPy~\cite{scipy}.  It
consists of a function that calls the Hamiltonian function supplied by
the user to evaluate, reorder and return its partial derivatives, computed using
second order central difference formulas.
%
%The step parameters $\delta\nu_i$ for
%the derivation formulas must be supplied by the user.
%This function is then
%given
%as an argument to SciPy's ODE solver \texttt{solve\_ivp}.
%
For greater simplicity of use, it is wrapped in a class (named \texttt{HGrad})
whose objects store, on creation, all the problem physical and numerical
details, like the Hamilton function and its parameters, the number of degrees of
freedom, the values of the discretization steps $\delta\nu_i$, etc.
\texttt{HGrad} objects are defined as callable with signature \texttt{objname(t,
$\nu_i$)}, returning a list containing the values of the rhs functions $G_i$ in
Hamilton's equations for given values of the dynamical variables of the
dynamical variables $\nu_i$ at given time \texttt{t}. In this way, the calling
syntax of \texttt{HGrad} objects is consistent with the interface of the ODE
solver we chose for our implementation (SciPy's function \texttt{solve\_ivp})
regarding the specification of the rhs $G_i$ functions, so that
\texttt{HGrad} objects can be used
just as if they were explicitly coded python functions.  The whole class
definition (including comments and Python doc strings) fits in a few dozen lines
of code and is very easy to use.  Our code is available for download as free
software at github~\cite{nevh:2020}.  The repository also includes a
\texttt{jupyter notebook} showcasing several example applications.

Listing~\ref{lst:1dlho} presents a particularly simple application example, the
one-dimensional harmonic oscillator. Most of the lines of code displayed deal
with general adminis\-trative details (initialization of physical and numerical
paramenters, definition of initial state, and so on), also needed for standard
numerical resolutions of Hamilton's equations. In Listing~\ref{lst:1dlho}, the
sections specific to our method are lines 6-8 (definition of the Hamiltonian
function), line 13 (initialization of the discretization steps for the
calculation of partial derivatives) and line 19 (definition of the
\texttt{HGrad} object \texttt{G}).

\begin{lstlisting}[caption= {Numerical solution of Hamiltonian's equations for
the one-dimensional linear harmonic oscillator. Note that the user only supplies
code for the system's Hamilton function in lines 7-9. The remaining code lines
define initial state, physical and numerical parameters.},
numbers=left, stepnumber=1, label={lst:1dlho}, language=python
]
from scipy.integrate import solve_ivp
import nevh
PI = 3.141592653589793

# Hamiltonian function
def H(t, s, k, m):
    x, p = s
    return 0.5 * k * x**2 + 0.5 * p**2 / m

# Initial state: off equilibrium position, at rest (x=1, p=0)
initial_state = [1.0, 0.0]
# Discretization step parameters for numerical derivatives
ds = [0.1, 0.1]
# Hamiltonian parameters. With k=4\pi^2, m=1, the period is 1
kc = 4 * PI**2
mc = 1.0

# Create the HGrad object, use scipy.solve_ivp to
G = nevh.HGrad(H, ds, k=kc, m=mc)

# solve Halmilton's eqs numerically from t=0 to t=3
trj = solve_ivp(G, [0.0, 3.0], initial_state)
\end{lstlisting}
Note that the \texttt{HGrad} object \texttt{G} is supplied to scipy's ODE solver
\texttt{solve\_ivp} just as if \texttt{G} was the name of a Python function for
evaluating the rhs of Hamilton's equations.

Scipy's ODE solver function \texttt{solve\_ivp} returns a composite structure
(named \texttt{trj} in Listing~\ref{lst:1dlho}) that stores information
on the numerical solution of the ODE, for example, sampled values of the
coordinates and momenta at different times in the interval \texttt{[tmin,
tmax]}.
%are stored at the arrays \texttt{trj.y[0]}, \texttt{trj.y[1]}, while the
%sampling times are stored in \texttt{trj.t}. The sampling frequency can be
%adjusted by the user, but here we just took the default behaviour of
%\texttt{solve\_ivp}, which sets that frequency accordind to precision
%requirements (that we also didn't bother to specify, because it wasn't necessary
%for convergence in this example).

The values of the discretization steps used for the numerical computation of the
partial derivatives of the Hamiltonian, are stored in array \texttt{ds} in
Listing~\ref{lst:1dlho} (at line 13).  In the example shown these values are
actually arbitrary, because the Hamiltonian of the linear harmonic oscilator is a
quadratic function of all the coordinates and conjugate momenta and the central
difference formulas yield exact estimates of the derivatives of quadratic
functions. However, in other problems, the values of the discretizations steps
may need to be more carefully considered.

Figure~\ref{fig:1dlho} displays the values of the  coordinates and momenta of
the numerical solution obtained (circles and squares, respectively) in arbitrary
units, together with plots of the analytical solutions for the same problem
(continuous and dashed lines).
\begin{figure}[htb]
  \begin{center}
    \includegraphics[width=0.6\linewidth]{figs/shm.png}
  \end{center}
  \caption{Plot of position (\texttt{x}) and momentum (\texttt{p}) for the
  one-dimensional harmonic oscillator problem of Listing~\ref{lst:1dlho}.
  }\label{fig:1dlho}
\end{figure}
The correctness of the numerical approximation, at least for pedagogical
purposes only, is manifest.
